Relationships in the extremes and their influence on competition and coexistence

### 2021-09-24 paper/Paper.Rnw
small changes to intro

%***DAN: Jasmin to edit the below as follows:
%1) Topic sentence: Another potentially important reason to quatify the importance of ATA to species coexistence is that there may
%be reason to believe that climate change can alter ATAs of environmental variables implicated in coexistence, as follows. 
%2) Just adapt below text, maybe try to simplify a bit.
%3) This para could just focus on the mechanics of how ATAs change for two variables measured at different times. 
%***DAN: Keeping this sentence so Jasmin can put it somwhere, not here.
%Given that storage effect depends on the relationship between E and C, we question the consequences of EC relationships characterized by ATA, the presence of which may be increasing as climate change brings about more extreme climatic events. 
%***DAN: I am also having a hard time identifying the topic sentence of this paragraph, can you tell me which one it is, please,
%in a comment right under this one?
% Extreme events can have important ecological impacts for storage effect and coexistence by introducing ATAs. 
An additional driving reason to quantify the significance of ATAs for species coexistence is the potential for climate change to alter ATAs of environmental variables required in coexistence, as follows. 
Among the various implications of climate change, the most potentially influential to ATAs in the environment is the increase in intensity, duration, extent, and frequency of extreme climatic events \citep{Jentsch_2007, Meehl_2004, Ummenhofer_2017}, i.e., heat-waves, floods, hurricanes, fires, etc. 
Temporally extensive extreme events can have important ecological impacts for
competing populations where species growth and the maintenance of biodiversity is dependent on phenologically separated environmental signals. A temporally extensive event can equalize the phenologically separated environmental signals, making them more correlated. For example, consider two precipitations signals, one in early spring and one in late summer, that are usually different. If a long flood occurs, it's likely that both signals will be close to equal in that year. If extreme floods occur in more years than not, then the two signals will become more correlated in higher precipitation values, a right-tailed ATA (figure \ref{fig:noise}C). Similarly, if extreme droughts occur in more years than not, then the signals will be more correlated in the lower values, a left-tailed ATA (figure \ref{fig:noise}A). The same can be said about spatially separated signals and more spatially extensive extreme events. ATA of environmental signals can then introduce or change ATAs in the fitness or traits of species that depend on the signals and potentially affect their coexistence.
%***DAN:
%1) Topic sentence: 
To make the above ideas more concrete, we review a relevant example from the literature that may be
influenced as previously described. 
We introduce a simplified example of an Arizona desert annual plant community from \cite{Chesson_2012}. 
%2) Make it concrete - name the system, name the organisms, name the seasons. 
The community follows the criteria required for temporal storage effects to operate. Different species dominate and have high germination rates depending on rainfall timing. For example, Baileya multiradiata dominates if sufficient rainfall is reached in the early germination season. Eriogonum abertianum dominates if sufficient rainfall is reached in the late germination season. Competition is influenced by resource consumption of seed-producing adult plants. Dormant seeds are unaffected by competition and can persist in the community in an unfavorable environment. When rain fall dynamics temporally fluctuate from year to year, relative abundance and dominance of the desert annual community also fluctuate year to year. The result is a relationship between the desert annuals' environmental responses (germination) that is symmetrically associated...

@@

Environmental fluctuations are directly impacted by climate change through changes in means and increased variance \citep{Hansen_2012}. While both changes can be attributed to extreme events, shifts in mean are more-so a result of gradual trend effects, and increased variance is a result of extreme event effects \citep{Jentsch_2007} .
Impacts on competition and coexistence of climate change have been more commonly researched via studying changes in the mean of environmental variables. 

@@

On the contrary, dominant species exclusion may be beneficial as rare species can increase in diversity \citep{Mariotte_2013}. Changes in environmental means can also increase fitness of a dominant species and in turn reduce community diversity, typical of invasive species \citep{Vitousek_1997}. 
As a result of changing means, range shifts can favor plant coexistence by introducing new herbivores\cite{Descombes_2020} or disfavor coexistence through increased competition \cite{Usinowicz_2021}.  

@@

Although, the body of literature on the impact of increased environmental variability has expanded, event-based studies accounted for only one-fifth of experimental climate change research as of 2006 \citep{Jentsch_2007}, the most commonly manipulated extremes being temperature and rainfall. Extreme drought and heat waves have shown to impact competitive output, rearranging species composition and dominance of grass-forb communities \citep{White_2000, White_2001, Hoover_2014}. However, complete exclusion of dominant species due to extreme events is uncommon \citep{Lloret_2012}. \cite{Fay_2003} experimentally manipulated rainfall for more frequent and intense precipitation while keeping constant the total volume of rain over Kansas tallgrass prairie and found that while some species had reduced fitness, overall community structure remained the same. For invasive species, increased climate and resource variability allowed for greater success of invasibility in terms of density and competitive equilibrium \citep{Parepa_2013}. 

@@

Although, increased variability of environmental fluctuations is a newer consideration for studying environmental changes, environmental variability has been analytically shown to be important for classical coexistence \citep{Chesson_1981, Warner_1985}. Moreover, by using MCT, increased variability has been shown to promote coexistence in plants communities, via the storage effect \citep{Adler_2006,  Angert_2009}. This present study has applications to extreme events and climate change as we consider relationships in their extremes, potentially caused by extreme events, and their implications for competition and coexistence. 
Though previous literature considers impacts of extreme events, change in means, and increased variability on species coexistence and competition and community dynamics in general (see above), they do not directly consider ATA between species response caused by environmental changes -- extreme relationships caused by extreme events (but see \cite{Ghosh_2021} for impact of ATA on community aggregate properties).

@@

%Move up the lottery model para from Theory to here, suture up the rift that this will produce in theory
To further demonstrate plausibility of ATA altering coexistence between species, we carried out simulations using a simple two-species model, the lottery model. 
Environmental variability has been classically shown to promote coexistence via storage mechanisms in a temporal lottery model of competition \citep{Chesson_1981, Warner_1985}, 
hence we move forward with this simple model to examine the effect of extreme relationships between environmental responses on the contribution of storage effects to coexistence. 
We introduce the lottery model and describe simulations based on it that clearly show that ATAs can influence 
species coexistence, though the details of this influence are the subject of the rest of the paper.
The lottery model is as follows.
Letting $N_i(t)$ denote the population density of species $i=1,2$ at time $t$, and defining $N = N_1(t)+N_2(t)$,
model equations are
\begin{equation}
N_i(t+1)=(1-\delta)N_i(t)+\delta N \frac{B_i(t)N_i(t)}{B_1(t)N_1(t)+B_2(t)N_2(t)}  \label{model_eq}
\end{equation}
for $i=1,2$. Here, $\delta$ is a mortality rate, and $B_i(t)$ is the fecundity of species $i$ at time $t$. 
The model postulates that individuals die at rate $\delta$ at each time step, and are replaced by juveniles 
in proportion to the reproductive outputs of the two species that year.
For each $i$, we assume the random variables $B_i(t)$ are independent and identically distributed (iid) through
time. We let $B_i = \exp(b_i)$, where $\left(b_1, b_2\right)$ is some bivariate random variable with the property 
that $b_i$ is normally distributed with mean $\mu_i$ and variance $\sigma^2$. 
However, we do \emph{not} assume that $\left(b_1, b_2\right)$ is a 
bivariate normal distribution. We will consider various distributions of $\left(b_1, b_2\right)$ with the above 
properties, corresponding to various tail association cases. Note that $N$ is constant through time.
%very succinct description of sims you do (mostly it's in the figure caption)
%1-2 sentences on the differences observed
We simulated two different communities up to \Sexpr{M} time steps to hopefully make it clear to the that ATA makes a difference for competition and coexistence between species (see Figure \ref{fig:pedagog} for more details). The only distinction between the two simulations is that one depends on asymmetrically tail associated temporal dynamics and one does not. Markedly, the simulation with ATA has longer periods of one-species dominance than the simulation without, implying less recoverability when rare and thus less coexistence \citep{Chesson_2000} when environmentally dependent variables are asymmetrically tail associated.

%Goal para
We have demonstrated that ATAs exist and are common in nature \citep{Ghosh_2020_AER}, and we have elucidated how ATAs between environmental response variables might arise via extreme events. ATA in environmental response directly feeds into an ATA effect on the EC covariance contribution to the storage effect mechanism of coexistence. Thus we seek to explore the consequences of ATAs on coexistence, competition, and the maintenance of biodiversity.
This paper has the following goals. 
G1) To augment the modern theory of coexistence (MTC) to show how ATA plays a role. 
The MTC decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATA contributes to these components. 
We here examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; spatial variation can be considered in future work.
G2) To explore, using a simple model, the relative magnitudes of the contributions to coexistence that come
from ATAs compared to other mechanisms. This is to help illuminate whether ATA contributions to coexistence 
are likely to be negligible or meaningfully large. 
%G3) To identify and define the ATA rescue or ATA exlusion set of trait space range for which species will coexist or become competitively excluded in our model. 
%***DAN: This is a good additional goal. We will either have to have already defined the "ATA rescue and ATA exlusion sets of trait space"
%or else readers won't understand this as it was written, above (now commented). These are complex concepts, so it may be better to 
%save formal definitions for later and reword this. Here's a try at rewording: 
G3) To explore, using our simple model, whether species can coexist with
ATAs but not without, or vice versa. This is to begin to illuminate whether ATA contributions to biodiversity maintenance 
may be negligible or meaninfully large.
G4) To provide an empirical example of species coexistence, and to highlight the role ATAs play in coexistence for the example.
In the Discussion, we revisit the potential that climatic changes will influence
the roles of ATAs in species coexistence and the maintenance of biodiversity.
Overall, our study presents a new mechanism of coexistence and biodiversity maintenance which may be influenced by climate change, 
and a means for understanding its importance for real systems. This is not a good ending sentence.

>>>>

%***DAN: Jasmin to edit the below as follows:
%1) Topic sentence: Another potentially important reason to quatify the importance of ATA to species coexistence is that there may
%be reason to believe that climate change can alter ATAs of environmental variables implicated in coexistence, as follows. 
%2) Just adapt below text, maybe try to simplify a bit.
%3) This para could just focus on the mechanics of how ATAs change for two variables measured at different times. 
%***DAN: Keeping this sentence so Jasmin can put it somwhere, not here.
%Given that storage effect depends on the relationship between E and C, we question the consequences of EC relationships characterized by ATA, the presence of which may be increasing as climate change brings about more extreme climatic events. 
%***DAN: I am also having a hard time identifying the topic sentence of this paragraph, can you tell me which one it is, please,
%in a comment right under this one?
% Extreme events can have important ecological impacts for storage effect and coexistence by introducing ATAs. 
An additional driving reason to quantify the significance of ATAs for species coexistence is the potential for climate change to alter ATAs of environmental variables required in coexistence, as follows. 
Among the various implications of climate change, the most potentially influential to ATAs in the environment is the increase in intensity, duration, extent, and frequency of extreme climatic events \citep{Jentsch_2007, Meehl_2004, Ummenhofer_2017}, i.e., heat-waves, floods, hurricanes, fires, etc. 
Temporally extensive extreme events can have important ecological impacts for
competing populations where species growth and the maintenance of biodiversity is dependent on
phenologically separated environmental signals, for instance, equalizing the signals, making them more correlated. 
Consider two precipitations signals, one in early spring and one in late summer, that are usually different. 
If a long flood occurs, it's likely that both signals will be close to equal in that year. 
If extreme floods occur in more years than not, then the two signals will become more correlated in higher precipitation values, a right-tailed ATA (figure \ref{fig:noise}C). 
Similarly, if extreme droughts occur in more years than not, then the signals will be more correlated in the lower values, a left-tailed ATA (figure \ref{fig:noise}A). 
The same can be said about spatially separated signals and more spatially extensive extreme events. 
ATA of environmental signals can then introduce or change ATAs in the fitness or traits of species that depend on the signals and potentially affect their coexistence.
%***DAN:
%1) Topic sentence:

To make the above ideas more concrete, we review a relevant example from the literature that may be
influenced as previously described. 
We introduce a simplified example of an Arizona desert annual plant community from \cite{Chesson_2012}. 
%2) Make it concrete - name the system, name the organisms, name the seasons. 
The community follows the criteria required for temporal storage effects to operate. 
Different species dominate and have high germination rates depending on rainfall timing. 
For example, Baileya multiradiata dominates if sufficient rainfall is reached in the early germination season. Eriogonum abertianum dominates if sufficient rainfall is reached in the late germination season. Competition is influenced by resource consumption of seed-producing adult plants. Dormant seeds are unaffected by competition and can persist in the community in an unfavorable environment. When rain fall dynamics temporally fluctuate from year to year, relative abundance and dominance of the desert annual community also fluctuate year to year. The result is a relationship between the desert annuals' environmental responses (germination) that is symmetrically associated...

@@

Environmental fluctuations are directly impacted by climate change through changes in means and increased variance \citep{Hansen_2012}. While both changes can be attributed to extreme events, shifts in mean are more-so a result of gradual trend effects, and increased variance is a result of extreme event effects \citep{Jentsch_2007}.
However, impacts on competition and coexistence of climate change have been more commonly researched via studying changes in the mean of environmental variables. 

@@

Alternatively, dominant species exclusion may be beneficial as rare species can increase in diversity \citep{Mariotte_2013}, while an increase in dominant species fitness can reduce community diversity, typical of invasive species \citep{Vitousek_1997}. 
As a product of changing means, range shifts can favor plant coexistence by introducing new herbivores \cite{Descombes_2020} or disfavor coexistence through increased competition \cite{Usinowicz_2021}.  

@@

Although, the body of literature on the impact of increased environmental variability has expanded, event-based studies accounted for only one-fifth of experimental climate change research as of 2006 \citep{Jentsch_2007}, the most commonly manipulated extremes being temperature and rainfall. Extreme drought and heat waves have shown to impact competitive output, rearranging species composition and dominance of grass-forb communities \citep{White_2000, White_2001, Hoover_2014}; although, complete exclusion of dominant species is uncommon \citep{Lloret_2012}. \cite{Fay_2003} experimentally manipulated rainfall for more frequent and intense precipitation while keeping constant the total volume of rain over Kansas tallgrass prairie and found that while some species had reduced fitness, overall community structure remained the same. For invasive species, increased climate and resource variability allowed for greater success of invasibility in terms of density and competitive equilibrium \citep{Parepa_2013}. 

@@

Although, increased variability of environmental fluctuations is a newer consideration for studying environmental changes, environmental variability has been analytically shown to be important for classical coexistence, i.e. MCT \citep{Chesson_1981, Warner_1985}. Moreover, by using MCT, increased variability has been shown to promote coexistence in plants communities, via the aforementioned storage effect \citep{Adler_2006,  Angert_2009}. This present study has applications to extreme events and climate change as we consider relationships in their extremes, potentially caused by extreme events, and their implications for competition and coexistence. 
Though previous literature considers impacts of extreme events, change in means, and increased variability on species coexistence and competition and community dynamics in general, they do not directly consider ATA between species response caused by environmental changes -- extreme relationships caused by extreme events (but see \cite{Ghosh_2021} for impact of ATA on community aggregate properties).

@@

%Move upl the lottery model para from Theory to here, suture up the rift that this will produce in theory
To further demonstrate plausibility of ATA altering coexistence between species, we carried out simulations using a simple two-species model, the lottery model. 
Environmental variability has been classically shown to promote coexistence via storage mechanisms in a temporal lottery model of competition \citep{Chesson_1981, Warner_1985}, 
hence we move forward with this simple model to examine the effect of extreme relationships between environmental responses on the contribution of storage effects to coexistence. 
We introduce the lottery model and describe simulations based on it that clearly show that ATAs can influence 
species coexistence, though the details of this influence are the subject of the rest of the paper.
The lottery model is as follows.
Letting $N_i(t)$ denote the population density of species $i=1,2$ at time $t$, and defining $N = N_1(t)+N_2(t)$,
model equations are
\begin{equation}
N_i(t+1)=(1-\delta)N_i(t)+\delta N \frac{B_i(t)N_i(t)}{B_1(t)N_1(t)+B_2(t)N_2(t)}  \label{model_eq}
\end{equation}
for $i=1,2$. Here, $\delta$ is a mortality rate, and $B_i(t)$ is the fecundity of species $i$ at time $t$. 
The model postulates that individuals die at rate $\delta$ at each time step, and are replaced by juveniles 
in proportion to the reproductive outputs of the two species that year.
For each $i$, we assume the random variables $B_i(t)$ are independent and identically distributed (iid) through
time. We let $B_i = \exp(b_i)$, where $\left(b_1, b_2\right)$ is some bivariate random variable with the property 
that $b_i$ is normally distributed with mean $\mu_i$ and variance $\sigma^2$. 
However, we do \emph{not} assume that $\left(b_1, b_2\right)$ is a 
bivariate normal distribution. We will consider various distributions of $\left(b_1, b_2\right)$ with the above 
properties, corresponding to various tail association cases. Note that $N$ is constant through time.
%very succinct description of sims you do (mostly it's in the figure caption)
%1-2 sentences on the differences observed
We simulated two different communities up to \Sexpr{M} time steps to hopefully make it clear to the that ATA makes a difference for competition and coexistence between species (see Figure \ref{fig:pedagog} for more details). The only distinction between the two simulations is that one depends on asymmetrically tail associated temporal dynamics and one does not. Markedly, the simulation with ATA has longer periods of one-species dominance than the simulation without, implying less recoverability when rare and thus less coexistence \citep{Chesson_2000} when environmentally dependent variables are asymmetrically tail associated.

%Goal para
We have demonstrated that ATAs exist and are common in nature \citep{Ghosh_2020_AER}, and we have elucidated how ATAs between environmental response variables might arise via extreme events. ATA in environmental response directly feeds into an ATA effect on the EC covariance contribution to the storage effect mechanism of coexistence. Thus we seek to explore the consequences of ATAs on coexistence, competition, and the maintenance of biodiversity.
This paper has the following goals. 
G1) To augment the modern theory of coexistence (MTC) to show how ATA plays a role. 
The MTC decomposes a measure of coexistence into components due to each of several mechanisms. 
We address how ATA contributes to these components. 
We here examine mechanisms of coexistence that apply in models exhibiting solely temporal variation; spatial variation can be considered in future work.
G2) To explore, using a simple model, the relative magnitudes of the contributions to coexistence that come
from ATAs compared to other mechanisms. This is to help illuminate whether ATA contributions to coexistence 
are likely to be negligible or meaningfully large. 
%G3) To identify and define the ATA rescue or ATA exlusion set of trait space range for which species will coexist or become competitively excluded in our model. 
%***DAN: This is a good additional goal. We will either have to have already defined the "ATA rescue and ATA exlusion sets of trait space"
%or else readers won't understand this as it was written, above (now commented). These are complex concepts, so it may be better to 
%save formal definitions for later and reword this. Here's a try at rewording: 
G3) To explore, using our simple model, whether species can coexist with
ATAs but not without, or vice versa. This is to begin to illuminate whether ATA contributions to biodiversity maintenance 
may be negligible or meaninfully large.
G4) To provide an empirical example of species coexistence, and to highlight the role ATAs play in coexistence for the example.
In the Discussion, we revisit the potential that climatic changes will influence
the roles of ATAs in species coexistence and the maintenance of biodiversity.
Our study presents a new mechanism of coexistence and biodiversity maintenance which may be influenced by climate change, 
and a means for understanding its importance for real systems. Need better ending sentence.
